---
title: "Lab 10: Generalized Linear Models I"
date: April 1, 2026
---

```{r echo = F, message = F, cache = FALSE, include = FALSE}
library(tidyverse)
library(brms)
library(ggplot2)
library(loo)
library(knitr)
library(formatR)

opts_chunk$set(out.width = '1\\linewidth', options(width=60))
opts_knit$set(global.par = TRUE)
opts_chunk$set(echo=T, comment = '', class.output='sh',
               fig.width=8, fig.height=5, fig.align = "center")
```

## Overview

In previous labs, we focused on linear regression models where the outcome was continuous and normally distributed. 

Today, we transition to **Generalized Linear Models (GLMs)**:

* **Generalization:** We use non-normal data distributions and link functions rather than the normal distribution and identity link.

* **Prior Specification:** The use of link functions changes how we approach prior predictive simulations.

* **Interpretation:** Instead of interpreting direct changes in $Y$, the parameters are now on a transformed scale, such as log-odds.


## Moving Beyond Linear Regression

Previously, we considered the relationship between GPA and study time.

\begin{align*}
  GPA_i & \sim \text{Normal}(\mu_i, \sigma) \\
  \mu_i & = \alpha + \beta \times \text{time}_i
\end{align*}

In `brms`, we specified this model using `family = gaussian(link = "identity")`. 

## Binary Outcomes

What if instead of predicting GPA itself, we wanted to model the probability of earning an "A" average or better (GPA > 3.5)? 

A normal distribution is no longer appropriate. Instead, we can use the **Bernoulli distribution** to describe the distribution of the data.

* Parameter $p$ represents the probability of success.

* The **logit** function relates $p$ to our predictor variable(s).

\begin{align*}
  \text{A}_i & \sim \text{Bernoulli}(p_i) \\
  \text{logit}(p_i) & = \alpha + \beta \times \text{time}_i
\end{align*}

## Linear vs. Logit Visual

Notice the y-axes!

```{r echo = FALSE}
f <- function(a, b, time) a + b*time
f_logit <- function(a, b, time) plogis(a + b*time)

p1 <- ggplot(data.frame(x = c(0,5), y = c(0,1)), aes(x = x, y = y)) +
  stat_function(fun = f, args = list(a = -1, b = .6)) +
  labs(title = "Identity Link", x = "Time", y = "Probability") +
  theme_minimal()

p2 <- ggplot(data.frame(x = c(0,5), y = c(0,1)), aes(x = x, y = y)) +
  stat_function(fun = f_logit, args = list(a = -3, b = 1)) +
  labs(title = "Logit Link", x = "Time", y = "Probability") +
  theme_minimal()

gridExtra::grid.arrange(p1, p2, nrow=1)
```


## Link Functions: Why Logit?

If we used an identity link ($p = \alpha + \beta \times \text{time}$), the model might predict probabilities below 0 or above 1. The **logit function** maps the linear predictor to a valid 0–1 range.


* **Intercept ($\alpha$):** `plogis(a)` provides the probability of a pass when study time is zero.

* **Slope ($\beta$):** The sign of $b$ controls whether the relationship between time and the probability is positive or negative.




## Prior Prediction for GLMs

Because the link function makes the relationship between parameters and the outcome non-linear, **prior predictive simulation** is essential for determining if our priors are reasonable.


We will use the `gpa_study_hours.csv` dataset where GPA $\ge$ 3.5 is coded as 1 ("A") and GPA < 3.5 is 0.

```{r message = FALSE}
dat <- read_csv("gpa_study_hours.csv")
dat %>% glimpse
```


In `brms`, logistic regression is specified with ` family = bernoulli(link = "logit")`

* Use `get_prior` to see how to input priors

```{r}
brm_form <- A ~ study_hours

get_prior(brm_form, data = dat, family = bernoulli(link = "logit"))
```

## Prior Prediction for GLMs

Let's start our simulations with a $N(0, 1)$ prior on both $a$ and $b$.

```{r results = "hide", cache = TRUE}
prior1 <- prior(normal(0, 1), class = "b") + 
          prior(normal(0, 1), class = "Intercept")

prior_sim1 <- brm(brm_form, data = dat, family = bernoulli(link = "logit"),
                  prior = prior1, sample_prior = "only")
```


## Prior Prediction for GLMs

For GLMS, `method = "posterior_epred"` plots results on the transformed "probability" scale.

```{r message = FALSE}
conditional_effects(prior_sim1, method = "posterior_epred")
```

## Prior Prediction for GLMs

This doesn't make sense! 

* We would expect an *increase* in study hours to *increase the probability of an A $\rightarrow$ let's increase the mean of the prior on $b$ to .25 and decrease the standard deviation to .5.

* The intercept corresponds to the probability of an A with **zero** hours of study, this should be lower than logit(0) = 0.5 $\rightarrow$ let's decrease the mean of the prior on $a$ to -.5. 

```{r message = FALSE, cache = TRUE}
prior2 <- prior(normal(.25, .5), class = "b") + 
          prior(normal(-.5, 1), class = "Intercept")

prior_sim2 <- brm(brm_form, data = dat, family = bernoulli(link = "logit"),
                  prior = prior2, sample_prior = "only")
```

## Prior Prediction for GLMs

Good enough?

```{r}
conditional_effects(prior_sim2, method = "posterior_epred")
```

## Fit the Logistic Regression Model

```{r results = "hide", cache = TRUE}
m1 <- brm(
  formula = brm_form, 
  data = dat,
  family = bernoulli(link = "logit"),
  prior = prior2)
```

## Results

We were wrong about the prior for the intercept. Grade inflation at work!

```{r}
print(m1)
```

## Results

```{r}
conditional_effects(m1, method = "posterior_epred")
```

## Binomial vs. Bernouli

Remember that the binomial distribution is a generalization of the bernoulli distribution

* Bernoulli only has binary outcomes with probability $p$

```{}
brm(outcome ~ predictors, family = bernoulli(link = "logit"))
```


* Binomial is the count of binary successes, each with probability $p$

    - `nsuccesses` is the total number of successes
    - `ntrials` is the total number of trials (potential successes), must be greater than the number of successes
    - `trials()` is special syntax in `brms`, not a variable in your data

```{}
brm(nsuccesses | trials(ntrials) ~ predictors, family = binomial(link = "logit"))
```


## Binomial Models

In lecture, we looked at time 1 injuries by sport type (individual vs. team). Let's repeat this analysis for time 2.

```{r}
injury <- read_csv("injury.csv")

d_bernoulli <- injury %>% 
  select(sport, sport_type, injured_2) %>%
  drop_na()

d_binomial <- d_bernoulli %>% group_by(sport) %>% 
  summarize(n = n(),
            ninj = sum(injured_2), 
            type = sport_type[1]) %>%
  mutate(sport = factor(sport), type = factor(type))
```


## Binomial Models

For this model, we want to treat both `sport` and `type` as index variables.

```{r}
brm_form2 <- bf(ninj | trials(n) ~ a + b,
                a ~ 0 + sport,
                b ~ 0 + type, nl = TRUE)

get_prior(brm_form2, d_binomial,
               family = binomial(link = "logit"))
```


## Binomial Models

Use the priors we used in lecture:

```{r results = "hide", cache = TRUE}
inj_mod <- brm(brm_form2, d_binomial,
               family = binomial(link = "logit"),
               prior = prior(normal(-2, 1), nlpar = "a") + 
                       prior(normal(0, 1), nlpar = "b"))
```

## Binomial Models

```{r}
print(inj_mod)
```
## Binomial Models

```{r}
conditional_effects(inj_mod, method = "posterior_epred")
```



## Activity


1. Use the `d_bernoulli` data that we defined earlier `family = bernoulli(link = "logit")` approach to fit the same model that we just fit using the binomial link. Compare the numeric and visual output from the two models.

2. In lecture we also learned about multinomial logistic regression. Using the code shown in the lecture slides as a reference, use the (unaggregated) `injury` data to predict `sport` from any other variables in the data set (e.g., training hours, stiffness). As time allows, try to also conduct prior predictive simulation. What do you find?

3. For the `conditional_effects()` function, the arguments `method = posterior_predict` and `method = posterior_linpred` function differently than `method = posterior_epred`. Test out these other arguments for the models that we fit today. What information does each plot convey?