---
title: "Lab 11: Generalized Linear Models II"
date: April 15, 2026
---


## Review

-   Last time we learned abou the Bernoulli and Binomial GLMs with with logit link function.

-   Today we will learn more about some more GLMs:

    -   Posisson
    -   Negative Binomial
    -   Zero-inflated
    -   Ordered categorical

## Poisson and Negative Binomial


-   The Poisson distribution can represent the distribution of non-negative integers. For example, when $\lambda = 3$:

```{r echo = F}
barplot(dpois(c(0:10), 3), xlab = "count", names.arg = c(0:10))
```

-   We first encountered Poisson as a way to model the number of meteors in a specific time interval (Lecture/Lab 2).

## Poisson and Negative Binomial

Recall that the Poisson distribution has one parameter $\lambda > 0$. The "log" link is appropriate for poisson regression because it transforms a non-negative value to the entire real number line. This ensures that $\hat{\lambda$ is a positive value.

\begin{align*}
y_i & \sim Pois(\lambda_i) \\
\log(\lambda_i) & = {\sf linear\,\,combination\,\,of\,\,predictors}
\end{align*}

To use the Poisson distribution with a log link in `brms`, we set:

```{r eval = F}
family(poisson(link = "log"))
```

## Poisson and Negative Binomial

However, the Poisson distribution is restrictive and may not be a good fit to overdispersed data, that is, data that are more variable than the model (e.g., the Poisson) expects.

The negative binomial distribution allows more variability than the Poisson distribution does.

## Poisson and Negative Binomial

The expected value and variance of the negative binomial value is

$$E(X) = \mu$$
$$var(X) = \mu + \frac{\mu^2}{\phi}$$ 

By estimating a second parameter $\phi$, we can flexibly account for overdispersion.


## Poisson and Negative Binomial

In R's `dnbinom` function, the `size` argument corresponds to $\phi$.

```{r echo = T}
barplot(dnbinom(c(0:5), size = 3, mu = 3))
```

## Poisson and Negative Binomial

As with the Poisson distribution, we typically use a log link for $\mu$.

\begin{align*}
y_i & \sim NegBin(\mu_i, \phi) \\
\log(\mu) & = {\sf linear\,\,combination\,\,of\,\,predictors} \\
\phi & \sim \_\_
\end{align*}

To use the negative binomial distribution with a log link in `brms`, we set:

```{r eval = FALSE}
family(negbinomial(link = "log"))
```

## Example Data

In business, an **interlock** occurs when an individual sits on the executive board or holds an executive position for **more than one company at the same time**.

The `Ornstein` data in the `carData` package provides the number of interlocks for 248 Canadian businesses in the 1970s. 

* Today, we will predict the number of interlocks from each business's `assets`

* `assets` is in millions of dollars and highly skewed. We will have a more stable model (i.e., better fit, better MCMC convergence) if we log-transform `assets` as a predictor.


## Example Data

```{r results = "hide"}
library(tidyverse)
library(brms)
data(Ornstein, package = "carData")
```

```{r}
#| class-output: "nowrap-code"
Ornstein %>% mutate(log_assets = log(assets)) %>% summary
```

## Poisson

First, let us look at the prior setup for the Poisson model:

```{r}
get_prior(interlocks ~ log(assets),  Ornstein,
          family = poisson(link = "log"))
```
## Poisson

\begin{align*}
{\sf interlock}_i & \sim Pois(\lambda_i) \\
\log(\lambda_i) & = a + b $\times \ln({\sf assets})
\end{align*}

* $\lambda$ is both the mean and variance of the outcome variable

We may have no idea where to start with prior predictive simulation. Why not start with $N(0, 1)$ on both the intercept and the slope?

```{r cache = TRUE, results = "hide"}
prior1 <- prior("normal(0, 1)", class = "Intercept") + 
          prior("normal(0, 1)", class = "b")
ppsim1 <- brm(interlocks ~ log(assets),  Ornstein,
          family = poisson(link = "log"),
          prior = prior1, sample_prior = "only")
```


## Poisson and Negative Binomial

These priors lead to a few draws with very high predictions (max # of interlocks is 107 in the data)

::: {.panel-tabset}

## No y axis limits

```{r}
conditional_effects(ppsim1, spaghetti = TRUE, ndraws = 100)
```


## Adding y axis limits

```{r}
plot(conditional_effects(ppsim1, spaghetti = TRUE, ndraws = 100), plot = FALSE)[[1]] + 
  lims(y = c(0, 107))
```

:::

## Poisson

Although these priors make some silly predictions, they may be okay for our purposes. The limited $y$ axis plot shows many plausible mean predictions.


The full model is:

\begin{align*}
{\sf interlock}_i & \sim Pois(\lambda_i) \\
\log(\lambda_i) & = a + b $\times \ln({\sf assets}) \\
a & \sim N(0, 1) \\
b & \sim N(0, 1)
\end{align*}


## Poisson

Let's fit the model!

```{r cache = TRUE, results = "hide"}
m1 <- brm(interlocks ~ log(assets), Ornstein, 
          family = poisson(link = "log"), prior = prior1)
```


## Poisson

It looks like our standard normal priors weren't too far off from the final estimates.

```{r}
summary(m1)
```

## Poisson

The Poisson model does an okay job at predicting the mean, but misses much of the variability in the data.

```{r}
plot(conditional_effects(m1), points = TRUE)
```

## Negative Binomial

The negative binomial allows for overdispersion in the data and may do a better job than the Poisson at predicting points. 

```{r}
get_prior(interlocks ~ log(assets), Ornstein, family = negbinomial(link = "log"))
```

* The negative binomial is similar to the Poisson except that it now includes the "shape" parameter $\phi$. **Smaller** values of $\phi$ will give **more variable** predictions relative to the Poisson.


## Negative Binomial

Let's try $\phi \sim Exponential(10)$. This roughly means that the expected variance is  10x as large as that for the Poisson.

```{r cache = TRUE, results = "hide"}
prior2 <- prior("normal(0, 1)", class = "Intercept") + 
          prior("normal(0, 1)", class = "b") + 
          prior("exponential(10)", class = "shape")
ppsim2 <- brm(interlocks ~ log(assets), Ornstein, family = negbinomial(link = "log"),
              prior = prior2, sample_prior = "only")
```


## Negative Binomial

Although these priors make some very large predictions, this may be an appropriately skeptical prior.

```{r}
conditional_effects(ppsim2, method = "posterior_predict")
```


## Negative Binomial

```{r cache = TRUE, results = "hide"}
m2 <- brm(interlocks ~ log(assets), Ornstein, 
          family = negbinomial(link = "log"), prior = prior2)
```

## Negative Binomial

```{r}
summary(m2)
```


## Negative Binomial

The negative binomial does a better job at capturing values for companies with high assets, but not as good a job for companies with fewer assets.

```{r}
plot(conditional_effects(m2), points = TRUE)
```


## Zero Inflation

Perhaps we should take into account **zero-inflation**, the fact that several companies have **no** interlocks.

```{r}
#| class-output: "nowrap-code"
get_prior(interlocks ~ log(assets), Ornstein, family = zero_inflated_negbinomial("log"))
```

Now we have one more parameter `zi` which represents the **proportion of zeroes**.


## Zero Inflation

The Beta distribution is an appropriate prior for `zi` since it bounded between 0 and 1. We might expect that fewer than half of companies have no interlocks, so maybe a Beta(1, 5) prior is appropriate?

```{r}
curve(dbeta(x, 1, 5))
```

```{r results = "hide", cache = TRUE}
prior3 <- prior2 + prior("beta(1, 5)", class = "zi")

m3 <- brm(interlocks ~ log(assets), Ornstein, 
          family = zero_inflated_negbinomial(link = "log"),
          prior = prior3)
```


## Zero Inflation

```{r}
summary(m3)
```

## Zero Inflation

```{r}
plot(conditional_effects(m3), points = TRUE)
```

## Activity 1

Does accounting for zero inflation improve our model? Use the `loo` to compare the three models that we fit. Which provides the best tradeoff between complexity and fit?




## Ordered Categorical Variables

We have learned how to outcome variables that are both continuous and categorical. In addition to those, ordered categorical variables are often used in psychology in the form of Likert scales.

When we use Likert scale (or other ordered categorical variable) as our outcome, we need to specify the family as:

```{r eval = F}
family = cumulative(link = "logit")
```

## Ordered Categorical Variables

The `WVS` dataset includes information surveyed from the 1995-1997 World Values Survey.

We are interested in predicting responses to `poverty`: "Do you think that what the government is doing for people in poverty in this country is..." (1) "Too Little", (2) "About Right", (3) "Too Much"

```{r}
#| class-output: "nowrap-code"
data(WVS, package = "carData")
WVS %>% summary
```


## Ordered Categorical Variables


-   $logit[P({\sf too\,little})] = Q_1 = a_1 + b_1 * {\sf predictor}$

-   $logit[P({\sf too\,little}) + P({\sf about\,right})] = Q_2 = a_2 + b * \sf{predictor}$

-   $P({\sf too\,much}) = 1 - P({\sf about\,right}) - P({\sf too\,little})$

## Ordered Categorical Variables

We want to predict `poverty` (already an ordered categorical variable) from `country` (Australia, Norway, Sweden, USA). For ordered categorical outcomes, `brms` does not allow us to use the index variable approach. Instead, it looks as if Australia is the default category and that the `b` coefficients correspond to the differences between each country and Australia.

```{r}
get_prior(poverty ~ country, WVS,
          family = cumulative(link = "logit"))
```

## Ordered Categorical Variables

Let's try using standard normal priors for all intercepts and slopes.

```{r cache = TRUE, results = "hide"}
prior4 <- prior("normal(0, 1)", class = "b") + 
          prior("normal(0, 1)", class = "Intercept")

ppsim4 <- brm(poverty ~ country, WVS,
          family = cumulative(link = "logit"),
          prior = prior4, sample_prior = "only")
```

## Ordered Categorical Variables

These priors give mean predictions at the middle response value. Unfortunately, they're a bit narrower for Australia than the other countries, but not by much.

```{r}
conditional_effects(ppsim4, method = "posterior_epred")
```

## Ordered Categorical Variables

Predictions of actual responses span the available range and are the same for each country.

```{r}
conditional_effects(ppsim4, method = "posterior_predict")
```


## Ordered Categorical Variables

Let's fit the model!

```{r cache = TRUE, results = "hide"}
m4 <- brm(poverty ~ country, WVS,
          family = cumulative(link = "logit"),
          prior = prior4)
```


## Ordered Categorical Variables

```{r}
#| class-output: "nowrap-code"
summary(m4)
```

## Ordered Categorical Variables

`categorical = TRUE` allows us to see the actual posterior distribution of predictions. Do these results surprise you?

```{r}
conditional_effects(m4, categorical = TRUE)
```


## Activity 2

1. Choose one of the models that we fit in today's lecture. Select one or two other variables available in that data set to add to your model as predictors.

2. What prior will you choose for the new predictor(s)? Consider whether it is advisable to rescale the variable(s).

3. Fit the model and interpret the numeric/visual output.

4. Compare the model fit to that of the original model. Can you improve fit over the previous model?

Bonus. Explore different ways of visualizing your model results. Look at the options available in `conditional_effects` and/or consult the internet.

