---
title: "Lab 1: Introduction"
date: January 14, 2026
---

## Introduction

-   Review syllabus and course policies

-   Installing/update `R` and `RStudio`

-   Verify that you can compile Quarto or Markdown files to PDF or Word

-   Install `ggplot2`

-   Install `rstan` and `brms`

## Syllabus and Lab Policies

-   Labs

    -   Computation/programming is a critical part of modern Bayesian statistics.
    -   Lab attendance is required for students registered for the class.

-   Lab activity

    -   Each lab has one activity which is due at the end of the lab.
    -   Lab activities are for participation credit.
    -   If you cannot attend the lab, contact me as soon as possible
    -   In the event of an excused absence, due dates for lab activities can be negotiated.

-   Office hour

    -   My office hours are on Zoom by appointment (link in syllabus)
    -   Contact me at xchen358@fordham.edu

## Download/Update R and RStudio

-   Installation

    -   For R: <https://cran.r-project.org/>

    -   For RStudio: <https://posit.co/download/rstudio-desktop/>

-   To make sure everything works properly in this class, it is best to regularly update R and R packages to latest version.

-   As of early January 2026, the current R version is 4.5.2 and current RStudio version is 2026.01.0+392

## Prepare R and RStudio

-   Open RStudio

-   You may also want to install a TeX distribution if you want to compile your homework assignments as pdfs (instead of Word documents). To do so, you can run this code:

```{r eval = FALSE}
install.packages("tinytex")
tinytex::install_tinytex()
```

You may be prompted to install "git" using command line developer tools or Rtools. If so, follow the instructions printed on your screen.

## Compile to PDF or Word

::: panel-tabset

## Quarto

Use Quarto if it works on your computer.

1.  Open a file: File $\rightarrow$ New File $\rightarrow$ Quarto Document

2.  Select “Document” in the leftmost panel.

3.  Select "PDF" or "Word" as the output format and click "Create".

4.  A template Quarto document will be opened in RStudio. Click "Render" or File $\rightarrow$ Render Document. Save the file first.

5.  If successful, a Word or PDF file should appear in the same folder in which you saved the file.

## R Markdown

R Markdown is older than Quarto. Only use this option if Quarto gives errors (e.g., Quarto files don't render on Windows computers with ARM processors).

1.  Open a new file: File $\rightarrow$ New File $\rightarrow$ R Markdown

2.  Select “Document” in the leftmost panel.

3.  Select "PDF" or "Word" as the output format and click "OK".

4.  A template R Markdown document will be opened in RStudio. Click "Knit" or File $\rightarrow$ Knit Document. Save the file first.

5.  If successful, a Word or PDF file should appear in the same folder in which you saved the file.
:::

## Introduction to `ggplot2`

Plotting is an essential component in this class. We will use 

-   `ggplot2` is a R package which is part of the `tidyverse`.

-   If you have already installed `tidyverse`, you can use `ggplot2`.

```{r eval = FALSE}
# install.packages("tidyverse")
library(tidyverse)
```

-   In this lab, we will focus on plotting with `ggplot2`. Even though it is not always the best solution, it is well-integrated with the other packages that we will us for Bayesian statistics and it is user-friendly. If you prefer, you may instead use basic R tools to make plots in the assignments.

## `stan` and `brms`

-   `stan` is a command-line program for statistical modeling that is currently the best for general-purpose Bayesian computation.

-   Most users interact with `stan` through integrations with programming languages like `R` or `python`. 

-   In this class, we will primarily interact with `stan` through the `brms` package in `R`. A major advantage of `brms` is that its syntax is similar to `lm()`.

```{r eval = FALSE}
# Ordinary regression:
lm(outcome ~ predictor1 + predictor2)

# Bayesian regression with brms:
brm(outcome ~ predictor1 + predictor2)
```

`brms` calls another `R` integration to `stan`: `rstan`.

## `rstan`

To install, follow the instructions [here](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started)

To verify your installation:

```{r eval = FALSE}
example(stan_model, package = "rstan", run.dontrun = TRUE)
```

You might see a long list of scary messages and it may take a long time. If successful, the last few lines should look like this:

```
Chain 4: Iteration: 1000 / 2000 [ 50%]  (Warmup)
Chain 4: Iteration: 1001 / 2000 [ 50%]  (Sampling)
Chain 4: Iteration: 1200 / 2000 [ 60%]  (Sampling)
Chain 4: Iteration: 1400 / 2000 [ 70%]  (Sampling)
Chain 4: Iteration: 1600 / 2000 [ 80%]  (Sampling)
Chain 4: Iteration: 1800 / 2000 [ 90%]  (Sampling)
Chain 4: Iteration: 2000 / 2000 [100%]  (Sampling)
Chain 4:
Chain 4:  Elapsed Time: 0.003 seconds (Warm-up)
Chain 4:                0.003 seconds (Sampling)
Chain 4:                0.006 seconds (Total)
Chain 4:
```

## `brms`

Remember: `brms` references `rstan` but is more user-friendly.

Run the following code to install `brms`. You may be prompted to install additional build tools (do this!). 

```{r eval = FALSE}
install.packages("brms")
```

Run the following code. If your `brms` installation was successful, the output should end with something similar to what is shown on the previous slide.

```{r eval = FALSE, cache = TRUE}
library(brms)
brm(zAge ~ zBase, data = epilepsy)
```

## Today's Activity

-   Install R, ggplot2 and `brms`.

-   Use the examples in this lab to verify correct installation.

-   To receive participation credit for today, verify with lab instructor that `brms` works successfully on your computer and that you can compile a document to Word or PDF using RStudio.

    - If needed, you can arrange to resolve installation issues during office hours.

-   It is extremely important that you successfully complete these steps to be able to participate in the rest of the semester. If you run into problems, reach out for help right away!
