---
title: "Lab 8: Model Comparison"
date: March 18, 2026
---

## Overview

<!-- need to focus this lab more on practical computational aspects -->

<!-- Fitting polynomial and/or interactive models in brms -->
<!-- Calculating psis -->
<!-- Moment matching -->
<!-- WAIC -->
<!-- loo_compare -->
<!-- model averaging -->

<!-- Model selection - weather data could be a good activity example -->


In previous labs, we learned how to use `brms` to set priors, fit models, and summarize results. 

Today we will learn how to compare the relative fit among models.

<!-- * We will also learn how to estimate polynomial regression and models with interactions. -->

First, load in the packages that we'll use today.

```{r}
library(tidyverse)
library(brms)
```

## Data Example

Let's return to the cycling data from earlier this semester and run that model:

* The example is old, but the data is new, download it again from this week's lab folder!

```{r cache = TRUE}
TDF <- read_csv("TDF.csv")
TDF<-TDF %>% 
  # mean-center height and convert to cm
  mutate (HeightC = (Height - mean(Height))*100)

brms_form <- Weight ~ HeightC

brms_prior<-prior(normal(66.5, 2), class = "Intercept") +
            prior(normal(1, .5), class = "b") +
            prior(exponential(.25), class = "sigma")

mod1 <- brm(brms_form,
            data = TDF,
            family = gaussian(link = "identity"),
            prior = brms_prior)
```


## Data Example

```{r}
summary(mod1)
```

## Log Likelihood

We can use `log_lik` to get the log density of the model for each iteration and each person.

* This is a measure of how well the data fit to the model.

```{r}
res_l <- log_lik(mod1)
```

The output is pointwise, meaning that each of the 4000 MCMC iterations and each of the 179 individuals in the data has its own contribution to the log density: 

```{r}
res_l %>% glimpse
```

## Out-of-Sample Deviance

The log pointwise density is a measure of goodness-of-fit **in the current sample**.

We usually care more about how the model performs **in new samples**.

We will learn about two methods to compare the relative performance of models in new samples:

1. `loo`: Approximate leave-one-out cross-validation via Pareto-smoothed importance sampling (PSIS)

2. `waic`: Balances in-sample prediction with a penalty term, similar to other information criteria

## Out-of-Sample Deviance

- Pareto-smoothed importance sampling (PSIS) is computationally efficient *approximation* to leave-one-out cross-validation.

- *Exact* leave-one-out cross-validation would involve fitting the model $N$ times, once for each of the $N$ rows in the data!

- `loo` is pointwise. We get one set of fit information for each row in our data:

```{r}
loo_res <- loo(mod1)
head(loo_res$pointwise)
```

## Out-of-Sample Deviance

If a data point has a high Pareto k value (e.g., > .7), then the loo approximation is not as accurate as we would like.

- None of our data points have Pareto k > .7.


```{r}
max(loo_res$pointwise[, 5])
```


## Out-of-Sample Deviance

Widely applicable information criterion (WAIC)

```{r}
waic_res <- waic(mod1)
waic_res
```

The `waic` row represents the information criterion we are looking for.

## Out-of-Sample Deviance

Like `loo`, `waic` is pointwise - this can be useful for understanding which indviduals contribute more/less to overall fit.

```{r}
head(waic_res$pointwise)
```

## Model Comparison

Back to the cycling data. We will fit new models that incorporate information about riders' specialties.

* `Leader`: riders who aim to win the entire 3-week Tour-de-France

* `Climber`: riders who focus on results for hilly and mountainous terrain

* `Sprinter`: riders who focus on results for flat terrain

* `All-rounder`: riders who have no particular specialty

```{r}
table(TDF$Specialty)
```

## Model Comparison

Let's add `Specialty` as an index variable intercept into our model.


```{r cache = TRUE}
brms_form <- Weight ~ 0 + Specialty + HeightC

brms_prior<-prior(normal(66.5, 2), class = "b") +
            prior(normal(1, .5), class = "b", coef = "HeightC") +
            prior(exponential(.25), class = "sigma")

mod2 <- brm(brms_form,
            data = TDF,
            family = gaussian(link = "identity"),
            prior = brms_prior)
```

## Model Comparison

```{r}
#| class-output: "nowrap-code"
summary(mod2)
```


## Model Comparison

```{r}
effects <- conditional_effects(mod2, effects = "HeightC", 
                         conditions = make_conditions(mod2, vars = "Specialty"), 
                         method = "posterior_linpred")
```

::: panel-tabset

## Multiple Plots

```{r}
plot(effects, facet_args = list(nrow = 1, ncol = 4))
```

## Single Plot

```{r}
ggplot(effects[[1]], aes(x = HeightC, color = Specialty, fill = Specialty)) + 
       geom_line(aes(y = estimate__)) + 
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = .2, linetype = 0)
```


:::

## Model Comparison

The previous model estimated only a separate intercept for each specialty. We may wish to consider a separate intercept and slope for each specialty:

```{r cache = TRUE}
brms_form <- Weight ~ 0 + Specialty + Specialty:HeightC

brms_prior<-prior(normal(66.5, 2), class = "b") +
            prior(normal(1, .5), class = "b", coef = "SpecialtyAllMrounder:HeightC") +
            prior(normal(1, .5), class = "b", coef = "SpecialtyClimber:HeightC") +
            prior(normal(1, .5), class = "b", coef = "SpecialtyLeader:HeightC") +
            prior(normal(1, .5), class = "b", coef = "SpecialtySprinter:HeightC") +
            prior(exponential(.25), class = "sigma")

mod3 <- brm(brms_form,
            data = TDF,
            family = gaussian(link = "identity"),
            prior = brms_prior)
```

## Model Comparison

```{r}
#| class-output: "nowrap-code"
summary(mod3)
```


## Model Comparison

```{r}
effects <- conditional_effects(mod3, effects = "HeightC", 
                  conditions = make_conditions(mod3, vars = "Specialty"))
```

::: panel-tabset

## Multiple Plots

```{r}
plot(effects, facet_args = list(nrow = 1, ncol = 4))
```



## Single Plot

```{r}
ggplot(effects[[1]], aes(x = HeightC, color = Specialty, fill = Specialty)) + 
       geom_line(aes(y = estimate__)) + 
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = .2, linetype = 0)
```

:::

## Model Comparison

We can compare the fit of our three models.

* Use `add_criterion` to compute the `loo` and `waic` 

```{r}
mod1 <- add_criterion(mod1, criterion = c("loo", "waic"))
```

```{r}
mod2 <- add_criterion(mod2, criterion = c("loo", "waic"))
```

```{}
Warning: 
1 (0.6%) p_waic estimates greater than 0.4. We recommend trying loo instead.
```

```{r}
mod3 <- add_criterion(mod3, criterion = c("loo", "waic"))
```
```{}
Warning: 
3 (1.7%) p_waic estimates greater than 0.4. We recommend trying loo instead.
```

There are some warnings for models 2 and 3. `loo` may be more reliable than `waic` for these models.


## Model Comparison

Now we can conduct model comparison. 

```{r}
loo_compare(mod1, mod2, mod3, criterion = "loo")
```

```{r}
loo_compare(mod1, mod2, mod3, criterion = "waic")
```

* Models 2 and 3 provide much better fit than model 1. 

* The fit of Model 2 is about 1.5 standard errors better than the fit of model 3.


In this example, `loo` and `waic` give almost exactly the same results, but they are slightly different:

```{r}
mod1$criteria$loo$estimates
mod1$criteria$waic$estimates
```

## Activity

Loed in the `weather.csv` data. In this activity, we will predict averag high temperature by year through linear and polynomial models.

```{r}
library(tidyverse)
library(brms)
temps <- read_csv("weather.csv")
```

Rescale years so that 1918 (The first year in this data set) is 0.

```{r}
temps <- temps %>% mutate(YEAR = YEAR - 1918)
```


For all models that you fit, use the following priors:

```{r}
temps_prior <- prior(normal(45, 5), class = "Intercept") + 
               prior(normal(0, 5), class = "b") + 
               prior(lognormal(2, 1), class = "sigma")
```


## Activity

Fit linear, quadratic, and cubic models using the following formulas

* `poly` creates a polynomial function

* `raw = TRUE` means to use ordinary polynomials (instead of [orthogonal polynomials](https://en.wikipedia.org/wiki/Orthogonal_polynomials))

```{r}
form_lin <- "Avg_High_Temp ~ YEAR"
form_quad <- "Avg_High_Temp ~ poly(YEAR, 2, raw = TRUE)"
form_cubic <- "Avg_High_Temp ~ poly(YEAR, 3, raw = TRUE)"
```

```{r echo = FALSE, eval = FALSE}
temps_mod1 <- brm(form_lin, temps, 
                 family = gaussian(link = "identity"),
                 prior = temps_prior)

temps_mod2 <- brm(form_quad, temps, 
                 family = gaussian(link = "identity"),
                 prior = temps_prior)

temps_mod3 <- brm(form_cubic, temps, 
                 family = gaussian(link = "identity"),
                 prior = temps_prior)
```

Review:

* Evaluate the **convergence** of each model. Do you think that the model summaries are reliable? 

<details>

<summary>Spoilers! Open me after evaluating convergence for yourself</summary>

You likely found that the cubic model did not fit well. It turns out that orthogonal polynomials help improve convergence by making the different polynomial terms uncorrelated with each other.

* Fit the cubic model again using the following model formula:

```{r}
form_cubic <- "Avg_High_Temp ~ poly(YEAR, 3, raw = FALSE)"
```


```{r echo = FALSE, eval = FALSE}
temps_mod3 <- brm(form_cubic, temps, 
                 family = gaussian(link = "identity"),
                 prior = temps_prior)
```

</details>

## Activity 

1. Compare the fit of the linear, quadratic, and cubic models using `loo` and `waic`. What do you conclude?

```{r echo = FALSE, eval = FALSE}
temps_mod1 <- add_criterion(temps_mod1, criterion = c("loo", "waic"))
temps_mod2 <- add_criterion(temps_mod2, criterion = c("loo", "waic"))
temps_mod3 <- add_criterion(temps_mod3, criterion = c("loo", "waic"))

loo_compare(temps_mod1, temps_mod2, temps_mod3, criterion = "loo")
loo_compare(temps_mod1, temps_mod2, temps_mod3, criterion = "waic")

# linear model fits best but the quadratic model is within a standard error
```

2. Plot the best fit curve and standard error for your quadratic model.

```{r echo = FALSE, eval = FALSE}
conditional_effects(temps_mod2, method = "posterior_linpred")
```

